\textcolor{black}{In-silico molecular transport via perivascular networks in the human intracranial space}

### 2025-06-27 paper/main.tex
change molecular enrichment with mol transport or tracer enrichment

\begin{abstract}
  The mechanisms of intracranial solute transport are fundamental to human brain health, with alterations often linked to disease and functional impairment, and with distinct opportunities for personalized diagnostics and treatment. However, our understanding of these mechanisms and their interplay remains incomplete, in part due to the complexity of integrating insights across scales, between species and from different modalities. Here, we combine mixed-dimensional modelling, multi-modal magnetic resonance images, and high performance computing to construct and explore a high-fidelity in-silico model of human intracranial molecular enrichment.  This model predicts the temporo-spatial spreading of a solute within an image-derived geometric representation of the subarachnoid space, ventricular system and brain parenchyma, including networks of surface perivascular spaces (PVSs). Our findings highlight the significant impact of cerebrospinal fluid (CSF) production and intracranial pulsatility on molecular enrichment following intrathecal tracer injection. We demonstrate that low-frequency vasomotion induces moderate CSF flow in surface PVS networks which substantially enhances tracer enrichment, and that impaired enrichment is a direct natural consequence of enlarged surface PVSs. This openly available technology platform thus provides an opportunity for integrating separate observations on diffusion in neuropil, vascular dynamics, intracranial pulsatility, CSF production, and efflux, and for exploring drug delivery and clearance in the human brain.
\end{abstract}
\begin{document}

@@

% What is the big picture problem and why is it important? 
The mechanisms underlying molecular transport within the intracranial
space are fundamental to human brain health and
function. Neurodegenerative diseases such as Alzheimer's, Parkinson's,
and Huntington's disease are all associated with abnormal accumulation
of protein aggregates together with alterations in transport and
clearance characteristics~\cite{rasmussen2018glymphatic,
  harrison2020impaired, eide2023plasma, liu2024glymphatic}. Moreover,
sleep and conversely sleep-deprivation play a definite yet enigmatic
role in modulating molecular enrichment and
clearance~\cite{xie2013sleep, eide2021sleep, eide2022altered,
  miao2024brain, hauglund2025norepinephrine}. In the last decade,
established theories have been challenged by new findings on molecular
movement and exchange~\cite{iliff2012paravascular,
  ringstad2017glymphatic, louveau2017understanding,
  proulx2021cerebrospinal, bohr2022glymphatic}, including substantial
variability between individuals and between patient
cohorts~\cite{ringstad2018brain, eide2021direction, eide2021impaired,
  eide2022altered}. These observations provide distinct opportunities
(and challenges) for personalized medicine e.g.~for tailored
intrathecal delivery of chemotherapies\cite{lohela2022glymphatic} such
as methodextrate in acute lymphoblastic leukemia
patients\cite{kadan2009comparison}, and for early diagnostics of
impaired brain clearance\cite{eide2021clinical, van2024human}. In
spite of their importance, our understanding of these mechanisms is
incomplete with open challenges and significant debate -- in part
relating to the translation of knowledge between scales, species,
experimental protocols, clinical cohorts, and individuals.

@@

% What are our research questions and how do we answer them
In this study, by leveraging geometric model reduction and
mixed-dimensional modelling~\cite{masri2024modelling}, structural
magnetic resonance (MR) images~\cite{hodneland2019new}, and high
performance computing, we introduce an integrated computational model
of intracranial molecular transport. Focusing on the interplay between
perivascular pathways, pulsatility and CSF flow dynamics, the model
predicts the temporal evolution and spatial distribution of a solute
concentration within a detailed geometric representation of the human
SAS and ventricular system, in networks of surface PVSs (i.e.~those
surrounding arteries and veins in the SAS), and in the brain
parenchyma. In terms of transport dynamics, we account for
heterogeneous diffusion, dispersive mixing induced by cardiac and
respiratory pulsatility in the CSF spaces and PVSs, convective fluid
flow driven by CSF production and peristaltic pumping, as well as
solute exchange and clearance across semi-permeable
membranes. Comparing with glymphatic MRI
studies~\cite{ringstad2017glymphatic, ringstad2018brain,
  watts2019measuring, eide2024functional}, the in-silico predictions
accurately represent molecular enrichment patterns, timing and
intercompartmental distributions. This open platform~\cite{ZENODO}
thus provides a technological opportunity for qualitatively and
quantitatively exploring key open questions relating to molecular
movement within the human brain environment such as the role of
pulsatility, perivascular pathways, structural compartmentalization,
or morphology.

@@

\subsection*{Reduced CSF pulsatility strongly shifts intracranial enrichment patterns}
\begin{figure}%[h!]
\centering 
\includegraphics[width = 0.95\textwidth]{figures/figure2.png}
\caption{\textbf{The balance between CSF pulsatility and production shapes molecular enrichment patterns.}
A) CSF spaces: the third and fourth ventricle (V3\&4), lateral ventricles (LV) and SAS; 
B) Boundaries and interfaces: the spinal subarachnoid space (SSAS), upper and lower arachnoid membrane (AM-U and AM-L), pia (Pia) and lateral ventricles (LV); 
C) Schematic of the CSF flow model; 
D) CSF flow induced by CSF production with outflow allowed across the
upper convexity (AM-U); 
}
\label{fig:csf}
\end{figure}
\begin{figure}
\ContinuedFloat
\caption{(\textbf{cont.})
E) Histograms of the production-induced CSF velocities in the SAS and third and fourth ventricle (V3 \& V4) in terms of the relative frequency of the mean flow speed in each computational cell weighted by its volume; 
F) CSF flow induced at peak systolic blood inflow; 
G) Cardiac-induced dispersion factor $R_c$;
H) CSF flow induced at peak respiratory expansion; 
I) Respiration-induced dispersion factor $R_r$, $D = (1 + R_c + R_r) D^{\rm Gad}$; J--M): Histograms of the cardiac-induced flow velocities (J),
cardiac-induced dispersion factors (K),
respiratory flow velocities (L), respiratory dispersion factor (M);
N--Q): sagittal, coronal and axial planes of tracer concentration after 12h for different model variations: baseline (N), no CSF production (O), low dispersion (P), and high dispersion (Q).
}
\end{figure}

@@

\subsection*{Enlarged PVSs in the SAS delay periarterial and intracranial molecular enrichment}

@@

%% Summary of findings
We have presented a high-fidelity in-silico model of molecular
enrichment and clearance in human intracranial spaces, enabling
tailored predictions of the influence of CSF space and (peri)vascular
morphology, physiological factors such as cardiac and respiratory
pulsatility, as well as of pathological conditions such as enlarged
PVSs in the SAS. A key observation is that the balance between CSF
production and intracranial pulsatility significantly shapes the
large-scale features of molecular enrichment, with e.g.~ventricular
tracer reflux after intrathecal injection in the absence of CSF
production, and preference towards anterior brain regions with reduced
pulsatility. Indeed, the simulated distribution patterns cover a
substantial range of the individual variations observed clinically
\cite{ringstad2018brain}. In terms of perivascular pathways, we find
that even moderate CSF flow on the order of 0.1--1.0 \textmu m/s in
the surface PVSs results in earlier tracer enrichment around major
cerebral arteries. This effect is more pronounced -- with up to a
three-hour difference in time of arrival -- with mean PVS flow rates
of approximately 10 \textmu m/s. Our models of CSF flow in
perivascular networks, based on first principles and asymptotic
analysis, predict net flow of such magnitudes and both antegrade and
retrograde flow within the perivascular network. The PVS may thus
function as highways facilitating rapid transport, even in the absence
of structural barriers, though a sharp concentration difference
between the PVS and surrounding tissues is unlikely without such
barriers. Finally, dilated PVS in the SAS will cause a substantial reduction in
cardiac- and vasomotion-driven flow velocities, which in turn impedes
perivascular transport.

@@

% Tracer enrichment predictions
While molecular enrichment of the brain and surrounding CSF spaces
have been extensively studied in animal models, especially in murine
models in connection with studies of the glymphatic system, there are
fewer reports of contrast enrichment in human subjects over 0--24h. We
therefore mainly compare our in-silico predictions with the series of
papers by Eide and colleagues~\cite{ringstad2017glymphatic,
  ringstad2018brain, eide2021sleep, eide2024functional} and Watts et
al~\cite{watts2019measuring}. Ringstad et al.~\cite{ringstad2018brain}
report of tracer enrichment in a centripetal pattern, primarily in
regions near large cerebral surface arteries, peaking in the CSF
spaces between 6 and 9 hours, with parenchymal tracer content still
increasing after 24 hours, but with large individual
differences. Watts et al~\cite{watts2019measuring} report of a similar
enrichment pattern and a concentration peak in the SAS of $\approx$0.5
mmol/l (relative to a total amount of 1.0 mmol) occuring after 10 to 15 hours. 
Our in-silico enrichment patterns agree with these observations, though with an earlier rather than later peak in the CSF spaces (2 hours after first tracer appearance at the craniocervical junction). In particular, we note that the substantial individual variation observed clinically is comparable to the variation in in-silico enrichment patterns associated with reduced CSF production or reduced intracranial pulsatility. We find that about 20\% of the tracer reaching the cranium enters the parenchyma, which aligns with the reported peak concentration values of $0.5\,$mmol/l in the SAS and $0.1\,$mmol/l in the parenchyma reported by Watts et al~\cite{watts2019measuring}.

@@

% Variations with sleep, exercise
Lifestyle factors such as sleep~\cite{xie2013sleep, miao2024brain,
  bojarskaite2023sleep, eide2021sleep, vinje2023human,
  hauglund2025norepinephrine, larsen2024sleep},
exercise~\cite{holstein2018voluntary, yildiz2022immediate}, and
alcohol intake~\cite{lundgaard2018beneficial} affect molecular
enrichment and clearance of the brain. In particular, pial arteries
display higher-amplitude low-frequency vasomotion during NREM and
intermediate sleep states, while REM sleep is associated with reduced
PVS width~\cite{bojarskaite2023sleep}. Peristaltic pumping is more
effective under both of these configurations, resulting in higher
estimated net PVS flow velocities~\cite{gjerde2023directional}. Even
higher PVS velocities would amplify the PVS flow effects simulated
here, with earlier arrival times along the PVSs, clear enrichment in
adjacent spaces, and accelerated molecular enrichment
overall. Moreover, sleep is linked with low-frequency oscillations in
human CSF flow~\cite{fultz2019coupled}. As we have shown here, the
level of dispersion in the CSF spaces strongly shapes molecular
enrichment patterns. While we, at baseline, account for the dispersion
induced by cardiac and respiratory pulsatility, the high dispersion
scenarios considered can be viewed as representative of additional
pulsatility induced by e.g.~neural waves~\cite{fultz2019coupled,
  williams2023neural} or other respiration patterns depending on
activity level. More broadly, we consider this as a key strength of
the current in-silico framework: allowing for integrating and
exploring the mechanistic macroscale implications of changes in
pulsatility at smaller spatial or temporal scales.
This approach of modeling pulsatility through an effective dispersion 
factor is a simplification that does not capture the full complexity
of transient fluid dynamics, such as non-uniform wall motion and ependymal
cilia within the ventricular system~\cite{siyahhan2014flow}, dynamic CSF
flow at the craniocervical junction, or the effect of microanatomical
features such as subarachnoid trabeculae on flow resistance and dispersion
in the SAS. However, our approach is supported by recent clinical 
observations. High-resolution MRI measurements by Hirschler et
al.~\cite{hirschler2024region} revealed that CSF mobility in regions
with high pulsatility is approximately tenfold greater than the 
self-diffusion of water, an enhancement factor consistent with 
the values computed in our model.

>>>>

\begin{abstract}
  The mechanisms of intracranial solute transport are fundamental to human brain health, with alterations often linked to disease and functional impairment, and with distinct opportunities for personalized diagnostics and treatment. However, our understanding of these mechanisms and their interplay remains incomplete, in part due to the complexity of integrating insights across scales, between species and from different modalities. Here, we combine mixed-dimensional modelling, multi-modal magnetic resonance images, and high performance computing to construct and explore a high-fidelity in-silico model of human intracranial molecular transport.  This model predicts the temporo-spatial spreading of a solute within an image-derived geometric representation of the subarachnoid space, ventricular system and brain parenchyma, including networks of surface perivascular spaces (PVSs). Our findings highlight the significant impact of cerebrospinal fluid (CSF) production and intracranial pulsatility on molecular transport following intrathecal tracer injection. We demonstrate that low-frequency vasomotion induces moderate CSF flow in surface PVS networks which substantially enhances tracer enrichment, and that impaired enrichment is a direct natural consequence of enlarged surface PVSs. This openly available technology platform thus provides an opportunity for integrating separate observations on diffusion in neuropil, vascular dynamics, intracranial pulsatility, CSF production, and efflux, and for exploring drug delivery and clearance in the human brain.
\end{abstract}
\begin{document}

@@

% What is the big picture problem and why is it important? 
The mechanisms underlying molecular transport within the intracranial
space are fundamental to human brain health and
function. Neurodegenerative diseases such as Alzheimer's, Parkinson's,
and Huntington's disease are all associated with abnormal accumulation
of protein aggregates together with alterations in transport and
clearance characteristics~\cite{rasmussen2018glymphatic,
  harrison2020impaired, eide2023plasma, liu2024glymphatic}. Moreover,
sleep and conversely sleep-deprivation play a definite yet enigmatic
role in modulating molecular transport and
clearance~\cite{xie2013sleep, eide2021sleep, eide2022altered,
  miao2024brain, hauglund2025norepinephrine}. In the last decade,
established theories have been challenged by new findings on molecular
movement and exchange~\cite{iliff2012paravascular,
  ringstad2017glymphatic, louveau2017understanding,
  proulx2021cerebrospinal, bohr2022glymphatic}, including substantial
variability between individuals and between patient
cohorts~\cite{ringstad2018brain, eide2021direction, eide2021impaired,
  eide2022altered}. These observations provide distinct opportunities
(and challenges) for personalized medicine e.g.~for tailored
intrathecal delivery of chemotherapies\cite{lohela2022glymphatic} such
as methodextrate in acute lymphoblastic leukemia
patients\cite{kadan2009comparison}, and for early diagnostics of
impaired brain clearance\cite{eide2021clinical, van2024human}. In
spite of their importance, our understanding of these mechanisms is
incomplete with open challenges and significant debate -- in part
relating to the translation of knowledge between scales, species,
experimental protocols, clinical cohorts, and individuals.

@@

% What are our research questions and how do we answer them
In this study, by leveraging geometric model reduction and
mixed-dimensional modelling~\cite{masri2024modelling}, structural
magnetic resonance (MR) images~\cite{hodneland2019new}, and high
performance computing, we introduce an integrated computational model
of intracranial molecular transport. Focusing on the interplay between
perivascular pathways, pulsatility and CSF flow dynamics, the model
predicts the temporal evolution and spatial distribution of a solute
concentration within a detailed geometric representation of the human
SAS and ventricular system, in networks of surface PVSs (i.e.~those
surrounding arteries and veins in the SAS), and in the brain
parenchyma. In terms of transport dynamics, we account for
heterogeneous diffusion, dispersive mixing induced by cardiac and
respiratory pulsatility in the CSF spaces and PVSs, convective fluid
flow driven by CSF production and peristaltic pumping, as well as
solute exchange and clearance across semi-permeable
membranes. Comparing with glymphatic MRI
studies~\cite{ringstad2017glymphatic, ringstad2018brain,
  watts2019measuring, eide2024functional}, the in-silico predictions
accurately represent tracer enrichment patterns, timing and
intercompartmental distributions. This open platform~\cite{ZENODO}
thus provides a technological opportunity for qualitatively and
quantitatively exploring key open questions relating to molecular
movement within the human brain environment such as the role of
pulsatility, perivascular pathways, structural compartmentalization,
or morphology.

@@

\subsection*{Reduced CSF pulsatility strongly shifts intracranial enrichment patterns}
\begin{figure}%[h!]
\centering 
\includegraphics[width = 0.95\textwidth]{figures/figure2.png}
\caption{\textbf{The balance between CSF pulsatility and production shapes tracer enrichment patterns.}
A) CSF spaces: the third and fourth ventricle (V3\&4), lateral ventricles (LV) and SAS; 
B) Boundaries and interfaces: the spinal subarachnoid space (SSAS), upper and lower arachnoid membrane (AM-U and AM-L), pia (Pia) and lateral ventricles (LV); 
C) Schematic of the CSF flow model; 
D) CSF flow induced by CSF production with outflow allowed across the
upper convexity (AM-U); 
}
\label{fig:csf}
\end{figure}
\begin{figure}
\ContinuedFloat
\caption{(\textbf{cont.})
E) Histograms of the production-induced CSF velocities in the SAS and third and fourth ventricle (V3 \& V4) in terms of the relative frequency of the mean flow speed in each computational cell weighted by its volume; 
F) CSF flow induced at peak systolic blood inflow; 
G) Cardiac-induced dispersion factor $R_c$;
H) CSF flow induced at peak respiratory expansion; 
I) Respiration-induced dispersion factor $R_r$, $D = (1 + R_c + R_r) D^{\rm Gad}$; J--M): Histograms of the cardiac-induced flow velocities (J),
cardiac-induced dispersion factors (K),
respiratory flow velocities (L), respiratory dispersion factor (M);
N--Q): sagittal, coronal and axial planes of tracer concentration after 12h for different model variations: baseline (N), no CSF production (O), low dispersion (P), and high dispersion (Q).
}
\end{figure}

@@

\subsection*{Enlarged PVSs in the SAS delay periarterial and intracranial tracer enrichment}

@@

%% Summary of findings
We have presented a high-fidelity in-silico model of molecular
enrichment and clearance in human intracranial spaces, enabling
tailored predictions of the influence of CSF space and (peri)vascular
morphology, physiological factors such as cardiac and respiratory
pulsatility, as well as of pathological conditions such as enlarged
PVSs in the SAS. A key observation is that the balance between CSF
production and intracranial pulsatility significantly shapes the
large-scale features of tracer enrichment, with e.g.~ventricular
tracer reflux after intrathecal injection in the absence of CSF
production, and preference towards anterior brain regions with reduced
pulsatility. Indeed, the simulated distribution patterns cover a
substantial range of the individual variations observed clinically
\cite{ringstad2018brain}. In terms of perivascular pathways, we find
that even moderate CSF flow on the order of 0.1--1.0 \textmu m/s in
the surface PVSs results in earlier tracer enrichment around major
cerebral arteries. This effect is more pronounced -- with up to a
three-hour difference in time of arrival -- with mean PVS flow rates
of approximately 10 \textmu m/s. Our models of CSF flow in
perivascular networks, based on first principles and asymptotic
analysis, predict net flow of such magnitudes and both antegrade and
retrograde flow within the perivascular network. The PVS may thus
function as highways facilitating rapid transport, even in the absence
of structural barriers, though a sharp concentration difference
between the PVS and surrounding tissues is unlikely without such
barriers. Finally, dilated PVS in the SAS will cause a substantial reduction in
cardiac- and vasomotion-driven flow velocities, which in turn impedes
perivascular transport.

@@

% Tracer enrichment predictions
While tracer enrichment of the brain and surrounding CSF spaces
have been extensively studied in animal models, especially in murine
models in connection with studies of the glymphatic system, there are
fewer reports of contrast enrichment in human subjects over 0--24h. We
therefore mainly compare our in-silico predictions with the series of
papers by Eide and colleagues~\cite{ringstad2017glymphatic,
  ringstad2018brain, eide2021sleep, eide2024functional} and Watts et
al~\cite{watts2019measuring}. Ringstad et al.~\cite{ringstad2018brain}
report of tracer enrichment in a centripetal pattern, primarily in
regions near large cerebral surface arteries, peaking in the CSF
spaces between 6 and 9 hours, with parenchymal tracer content still
increasing after 24 hours, but with large individual
differences. Watts et al~\cite{watts2019measuring} report of a similar
enrichment pattern and a concentration peak in the SAS of $\approx$0.5
mmol/l (relative to a total amount of 1.0 mmol) occuring after 10 to 15 hours. 
Our in-silico enrichment patterns agree with these observations, though with an earlier rather than later peak in the CSF spaces (2 hours after first tracer appearance at the craniocervical junction). In particular, we note that the substantial individual variation observed clinically is comparable to the variation in in-silico enrichment patterns associated with reduced CSF production or reduced intracranial pulsatility. We find that about 20\% of the tracer reaching the cranium enters the parenchyma, which aligns with the reported peak concentration values of $0.5\,$mmol/l in the SAS and $0.1\,$mmol/l in the parenchyma reported by Watts et al~\cite{watts2019measuring}.

@@

% Variations with sleep, exercise
Lifestyle factors such as sleep~\cite{xie2013sleep, miao2024brain,
  bojarskaite2023sleep, eide2021sleep, vinje2023human,
  hauglund2025norepinephrine, larsen2024sleep},
exercise~\cite{holstein2018voluntary, yildiz2022immediate}, and
alcohol intake~\cite{lundgaard2018beneficial} affect molecular
enrichment and clearance of the brain. In particular, pial arteries
display higher-amplitude low-frequency vasomotion during NREM and
intermediate sleep states, while REM sleep is associated with reduced
PVS width~\cite{bojarskaite2023sleep}. Peristaltic pumping is more
effective under both of these configurations, resulting in higher
estimated net PVS flow velocities~\cite{gjerde2023directional}. Even
higher PVS velocities would amplify the PVS flow effects simulated
here, with earlier arrival times along the PVSs, clear enrichment in
adjacent spaces, and accelerated tracer enrichment
overall. Moreover, sleep is linked with low-frequency oscillations in
human CSF flow~\cite{fultz2019coupled}. As we have shown here, the
level of dispersion in the CSF spaces strongly shapes molecular
enrichment patterns. While we, at baseline, account for the dispersion
induced by cardiac and respiratory pulsatility, the high dispersion
scenarios considered can be viewed as representative of additional
pulsatility induced by e.g.~neural waves~\cite{fultz2019coupled,
  williams2023neural} or other respiration patterns depending on
activity level. More broadly, we consider this as a key strength of
the current in-silico framework: allowing for integrating and
exploring the mechanistic macroscale implications of changes in
pulsatility at smaller spatial or temporal scales.
This approach of modeling pulsatility through an effective dispersion 
factor is a simplification that does not capture the full complexity
of transient fluid dynamics, such as non-uniform wall motion and ependymal
cilia within the ventricular system~\cite{siyahhan2014flow}, dynamic CSF
flow at the craniocervical junction, or the effect of microanatomical
features such as subarachnoid trabeculae on flow resistance and dispersion
in the SAS. However, our approach is supported by recent clinical 
observations. High-resolution MRI measurements by Hirschler et
al.~\cite{hirschler2024region} revealed that CSF mobility in regions
with high pulsatility is approximately tenfold greater than the 
self-diffusion of water, an enhancement factor consistent with 
the values computed in our model.
